---
title: "Indices of deprivation: which is the worst local authority?"
author: "Paul Bradshaw"
date: "23/09/2019"
output: html_document
---

# Indices of deprivation: which is the worst local authority?

This notebook details the process of fetching and cleaning indices of deprivation (IoD) and analysing the local authority-level data.

***TO DO: ADD LOOKUP FROM OLD LA CODES TO 2015 ONES***

## What we need to do

We need to:

* Fetch the spreadsheets
* Clean where needed, e.g. making headings consistent
* Combine the data
* Analyse to see which are worst, which changed the most
* Export the results for visualisation

***CHANGE THE FIRST LINE OF CODE TO THE 2019 DATA URL ONCE IT IS LIVE***

```{r store urls}
#CHANGE THIS TO THE 2019 URL ONCE LIVE
dataurl19.la <- "https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/464465/File_11_ID_2015_Upper-tier_Local_Authority_Summaries.xlsx"
#Store both URLs as strings
dataurl15.la <- "https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/464465/File_11_ID_2015_Upper-tier_Local_Authority_Summaries.xlsx"
dataurl10.la <- "https://www.gov.uk/government/uploads/system/uploads/attachment_data/file/6884/1871689.xls"
```

Now to fetch those we need the `rio` library:

```{r install rio}
#Install rio if needed
if(!require("rio")){
  install.packages("rio")
}
#Activate rio
library(rio)
```

And fetch the URL:

```{r import from urls}
#Fetch the second sheet of that spreadsheet, and store
data19.la <- rio::import(dataurl19.la, sheet = 2)
data15.la <- rio::import(dataurl15.la, sheet = 2)
data10.la <- rio::import(dataurl10.la, sheet = 2)
```
The datasets differ slightly in terms of columns, but the first column in each has the LSOA name, and we need to make sure we can match on that. So we change the name of that column:

```{r rename la col}
#Change col 1
colnames(data19.la)[1] <- "lacode"
colnames(data15.la)[1] <- "lacode"
colnames(data10.la)[1] <- "lacode"
#LA names
colnames(data19.la)[2] <- "laname"
colnames(data15.la)[2] <- "laname"
colnames(data10.la)[2] <- "laname"
#IMD extent
colnames(data19.la)[9] <- "imdextent"
colnames(data15.la)[9] <- "imdextent"
colnames(data10.la)[5] <- "imdextent"
#IMD extent rank
colnames(data19.la)[10] <- "imdextentrank"
colnames(data15.la)[10] <- "imdextentrank"
colnames(data10.la)[6] <- "imdextentrank"
```


## Combining the data 

Now let's combine those:

```{r combine data}
library(dplyr)
#The 2010 data has different codes and names are not the same as 2015, so we don't join
#Then left join
ladata15to19 <- dplyr::left_join(data19.la, data15.la, by = "lacode")
```


## Analysis: what are the 10 most local authorities with the highest ranking extent?

```{r install sqldf}
if(!require("sqldf")){
  install_formats("sqldf")
}
library(sqldf)
```

Show the 10 lowest ranked LSOAs:

```{r 10 lowest ranked}
#Show the 10 lowest ranked LAs
#Names with . in them need to be in `` to avoid it being parsed as a table name
latop10 <- sqldf::sqldf("SELECT lacode, `laname.x`, `imdextentrank.x` AS rank15, `imdextentrank.x` AS rank19, `imdextentrank.y` - `imdextentrank.x` AS change1519 FROM ladata15to19
             ORDER BY rank19 ASC
             LIMIT 10")
print(latop10)
```

We can repeat this to see those with the biggest changes:

```{r 10 lowest ranked}
#A rank then of 10, minus a rank.y of 10 = a change of -9, in other words they have 'dropped' from 10th to 1st worst. The lower the number, the bigger the drop, so we order descending:
larisers <- sqldf::sqldf("SELECT lacode, `laname.x`, `imdextentrank.x` AS rank15, `imdextentrank.x` AS rank19, `imdextentrank.y` - `imdextentrank.x` AS change1519 FROM ladata15to19
             ORDER BY change1519 DESC
             LIMIT 10")
print(larisers)
```


And the biggest changes in the other direction:

```{r 10 lowest ranked}
# Reverse the order to get biggest fallers
ladroppers <- sqldf::sqldf("SELECT lacode, `laname.x`, `imdextentrank.x` AS rank15, `imdextentrank.x` AS rank19, `imdextentrank.y` - `imdextentrank.x` AS change1519 FROM ladata15to19
             ORDER BY change1519 ASC
             LIMIT 10")
print(ladroppers)
```

## Export the results

Now export as a spreadsheet:

```{r export csv}
write.csv(latop10, "latop10.csv")
write.csv(ladroppers,"ladroppers.csv")
write.csv(larisers,"larisers.csv")
```



