---
title: "Indices of deprivation: fetching and cleaning"
author: "Paul Bradshaw"
date: "23/09/2019"
output: html_document
---

# Indices of deprivation: fetching and cleaning

This notebook details the process of fetching and cleaning indices of deprivation (IoD) and analysing score-level data.

The overall IoD ranking is based on a score given to each LSOA. That score is calculated based on a series of other scores, for aspects of deprivation ranging from employment to living environment.

Some of these scores are meaningful and comparable - such as the % of people "involuntarily excluded from the labour market" - while others are less so, such as a nunmber between 4 and -4 to indicate living environment.

The *overall* rankings based on those scores are provided in a main spreadsheet, but the detail of the scores are elsewhere. In 2015 the individual scores were provided in a separate spreadsheet, [File 5: scores for the indices of deprivation](https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/467768/File_5_ID_2015_Scores_for_the_Indices_of_Deprivation.xlsx). In 2010, each 'domain' (e.g. housing) was published in its own separate spreadsheet.


## What we need to do

We need to:

* Fetch the spreadsheet
* Clean it where needed, e.g. making headings consistent
* Combine the data
* Analyse to see which aspects have changed
* Export the results for visualisation


## Fetching data from the URLs

First, we store the URLs for the spreadsheets with individual scores.

```{r store scores URL}
#CHANGE THIS URL WHEN 2019 IS PUBLISHED
scoresurl19 <- "https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/467768/File_5_ID_2015_Scores_for_the_Indices_of_Deprivation.xlsx"

#2015 scores
scoresurl15 <- "https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/467768/File_5_ID_2015_Scores_for_the_Indices_of_Deprivation.xlsx"
#This is just income scores
incomeurl10 <- "https://www.gov.uk/government/uploads/system/uploads/attachment_data/file/6873/1871528.xls"
```

Then we import from them:

```{r import data from url}
#Fetch the second sheet of that spreadsheet, and store
scores19 <- rio::import(scoresurl19, sheet = 2)
scores15 <- rio::import(scoresurl15, sheet = 2)
scores10.income <- rio::import(incomeurl10, sheet = 2)
```

## Clean the data

Now we rename the columns so that they are consistent

```{r rename la col}
#Change col 1
colnames(scores19)[1] <- "lsoacode"
colnames(scores15)[1] <- "lsoacode"
colnames(scores10.income)[1] <- "lsoacode"

colnames(scores19)[2] <- "lsoaname"
colnames(scores15)[2] <- "lsoaname"

colnames(scores19)[3] <- "lacode"
colnames(scores15)[3] <- "lacode"
colnames(scores10.income)[2] <- "lacode"

colnames(scores19)[4] <- "laname"
colnames(scores15)[4] <- "laname"
colnames(scores10.income)[3] <- "laname"
```

We also need to name columns so the year is obvious:

```{r more renaming}
colnames(scores19)[6] <- "incomescore19"
colnames(scores15)[6] <- "incomescore15"
colnames(scores10.income)[6] <- "incomescore10"

colnames(scores19)[7] <- "employscore19"
colnames(scores15)[7] <- "employscore15"

colnames(scores19)[13] <- "IDACIrate19"
colnames(scores15)[13] <- "IDACIrate15"
colnames(scores19)[14] <- "IDAOPIrate19"
colnames(scores15)[14] <- "IDAOPIrate15"

```



## Combining the data 

Now let's combine those:

```{r combine data}
library(dplyr)
#We use a right join because the 2015 data has more rows
scores10to19 <- dplyr::right_join(scores10.income, scores15, by = "lsoacode")
#Then left join
scores10to19 <- dplyr::left_join(scores10to19, scores19, by = "lsoacode")
```


## Analysis: what are the 10 LSOAs with the worst income rate?

```{r install sqldf}
if(!require("sqldf")){
  install_formats("sqldf")
}
library(sqldf)
```

Show the 10 LSOAs with the highest income rates:

```{r 10 lowest ranked}
#Show the 10 lowest ranked LAs
#Names with . in them need to be in `` to avoid it being parsed as a table name
incometop10 <- sqldf::sqldf("SELECT lsoacode, `lsoaname.x`, incomescore10, incomescore15, incomescore19 FROM scores10to19
             ORDER BY incomescore19 DESC
             LIMIT 10")
print(incometop10)
```

Show the LSOAs where that rate has changed the most:

```{r 10 biggest drops}

incomedrops <- sqldf::sqldf("SELECT lsoacode, `lsoaname.x`,  incomescore10, incomescore15, incomescore19, `incomescore19` - `incomescore15` AS change15to19 FROM scores10to19
             ORDER BY change15to19 DESC
             LIMIT 10")
print(incomedrops)
```

```{r 10 biggest drops 2}
#We add a WHERE filter to remove NA entries for 2010
incomerises <- sqldf::sqldf("SELECT lsoacode, `lsoaname.x`,  incomescore10, incomescore15, incomescore19, `incomescore19` - `incomescore15` AS change15to19 FROM scores10to19
WHERE `incomescore10` != 'NA'
             ORDER BY change15to19 ASC
             LIMIT 10")
print(incomerises)
```


## Analysis: what are the 10 LSOAs with the worst employment rate?



```{r 10 lowest ranked}
#Show the 10 lowest ranked LAs
#Names with . in them need to be in `` to avoid it being parsed as a table name
emptop10 <- sqldf::sqldf("SELECT lsoacode, `lsoaname.x`, employscore15, employscore19 FROM scores10to19
             ORDER BY employscore19 DESC
             LIMIT 10")
print(emptop10)
```

Show the LSOAs where that rate has changed the most:

```{r 10 biggest drops}

empdrops <- sqldf::sqldf("SELECT lsoacode, `lsoaname.x`, employscore15, employscore19, `employscore19` - `employscore15` AS change15to19 FROM scores10to19
             ORDER BY change15to19 DESC
             LIMIT 10")
print(empdrops)
```

```{r 10 biggest drops}
emprises <- sqldf::sqldf("SELECT lsoacode, `lsoaname.x`, employscore15, employscore19, `employscore19` - `employscore15` AS change15to19 FROM scores10to19
             ORDER BY change15to19 ASC
             LIMIT 10")
print(emprises)
```

## Analysis: what are the 10 LSOAs with the worst Income Deprivation Affecting Children Index (IDACI) and Income Deprivation Affecting Older People (IDAOPI) rates?


```{r 10 lowest ranked}
#Show the 10 lowest ranked LAs
#Names with . in them need to be in `` to avoid it being parsed as a table name
idacitop10 <- sqldf::sqldf("SELECT lsoacode, `lsoaname.x`, IDACIrate15, IDACIrate19 FROM scores10to19
             ORDER BY employscore19 DESC
             LIMIT 10")
print(idacitop10)
```

Show the LSOAs where that rate has changed the most:

```{r 10 biggest drops}

idacidrops <- sqldf::sqldf("SELECT lsoacode, `lsoaname.x`, IDACIrate15, IDACIrate19, `IDACIrate19` - `IDACIrate15` AS change15to19 FROM scores10to19
             ORDER BY change15to19 DESC
             LIMIT 10")
print(idacidrops)
```

```{r 10 biggest drops}
idacirises <- sqldf::sqldf("SELECT lsoacode, `lsoaname.x`, IDACIrate15, IDACIrate19, `IDACIrate19` - `IDACIrate15` AS change15to19 FROM scores10to19
             ORDER BY change15to19 ASC
             LIMIT 10")
print(idacirises)
```

### IDAOPI 

```{r 10 lowest ranked}
#Show the 10 lowest ranked LAs
#Names with . in them need to be in `` to avoid it being parsed as a table name
idaopitop10 <- sqldf::sqldf("SELECT lsoacode, `lsoaname.x`, IDAOPIrate15, IDAOPIrate19 FROM scores10to19
             ORDER BY employscore19 DESC
             LIMIT 10")
print(idaopitop10)
```

Show the LSOAs where that rate has changed the most:

```{r 10 biggest drops}

idaopidrops <- sqldf::sqldf("SELECT lsoacode, `lsoaname.x`, IDAOPIrate15, IDAOPIrate19, `IDAOPIrate19` - `IDAOPIrate15` AS change15to19 FROM scores10to19
             ORDER BY change15to19 DESC
             LIMIT 10")
print(idaopidrops)
```

```{r 10 biggest drops}
idaopirises <- sqldf::sqldf("SELECT lsoacode, `lsoaname.x`, IDAOPIrate15, IDAOPIrate19, `IDAOPIrate19` - `IDAOPIrate15` AS change15to19 FROM scores10to19
             ORDER BY change15to19 ASC
             LIMIT 10")
print(idaopirises)
```

## Export the results

Now export as a spreadsheet:

```{r export csv}
write.csv(incometop10, "incometop10.csv")
write.csv(incomedrops,"incomedrops.csv")
write.csv(incomerises,"incomerises.csv")

write.csv(emptop10, "emptop10.csv")
write.csv(empdrops,"empdrops.csv")
write.csv(emprises,"emprises.csv")

write.csv(idacitop10, "idacitop10.csv")
write.csv(idacidrops, "idacidrops.csv")
write.csv(idacirises, "idacirises.csv")

write.csv(idaopitop10, "idaopitop10.csv")
write.csv(idaopidrops, "idaopidrops.csv")
write.csv(idaopirises, "idaopirises.csv")

```



