---
title: "Indices of deprivation: how has England changed? (detailed scores)"
author: "Paul Bradshaw"
date: "23/09/2019"
output: html_document
---

# Indices of deprivation: fetching and cleaning

This notebook details the process of fetching and cleaning indices of deprivation (IoD) and analysing score-level data.

The overall IoD ranking is based on a score given to each LSOA. That score is calculated based on a series of other scores, for aspects of deprivation ranging from employment to living environment.

Some of these scores are meaningful and comparable - such as the % of people "involuntarily excluded from the labour market" - while others are less so, such as a nunmber between 4 and -4 to indicate living environment.

The *overall* rankings based on those scores are provided in a main spreadsheet, but the detail of the scores are elsewhere. In 2015 the individual scores were provided in a separate spreadsheet, [File 5: scores for the indices of deprivation](https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/467768/File_5_ID_2015_Scores_for_the_Indices_of_Deprivation.xlsx). In 2010, each 'domain' (e.g. housing) was published in its own separate spreadsheet.


## What we need to do

We need to:

* Fetch the spreadsheet
* Clean it where needed, e.g. making headings consistent
* Combine the data
* Analyse to see which aspects have changed
* Export the results for visualisation


## Fetching data from the URLs

First, we store the URLs for the spreadsheets with individual scores.

```{r store scores URL}
#CHANGE THIS URL WHEN 2019 IS PUBLISHED
scoresurl19 <- "https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/467768/File_5_ID_2015_Scores_for_the_Indices_of_Deprivation.xlsx"

#2015 scores
scoresurl15 <- "https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/467768/File_5_ID_2015_Scores_for_the_Indices_of_Deprivation.xlsx"
#This is just income scores
incomeurl10 <- "https://www.gov.uk/government/uploads/system/uploads/attachment_data/file/6873/1871528.xls"
```

Then we import from them:

```{r import data from url}
#Fetch the second sheet of that spreadsheet, and store
scores19 <- rio::import(scoresurl19, sheet = 2)
scores15 <- rio::import(scoresurl15, sheet = 2)
scores10.income <- rio::import(incomeurl10, sheet = 2)
```

## Clean the data

Now we rename the columns so that they are consistent

```{r rename la col}
#Change col 1
colnames(scores19)[1] <- "lsoacode"
colnames(scores15)[1] <- "lsoacode"
colnames(scores10.income)[1] <- "lsoacode"

colnames(scores19)[2] <- "lsoaname"
colnames(scores15)[2] <- "lsoaname"

colnames(scores19)[3] <- "lacode"
colnames(scores15)[3] <- "lacode"
colnames(scores10.income)[2] <- "lacode"

colnames(scores19)[4] <- "laname"
colnames(scores15)[4] <- "laname"
colnames(scores10.income)[3] <- "laname"

colnames(scores19)[6] <- "incomescore19"
colnames(scores15)[6] <- "incomescore15"
colnames(scores10.income)[6] <- "incomescore10"

colnames(scores19)[5] <- "imdscore19"
colnames(scores15)[5] <- "imdscore15"

```




## Combining the data 

Now let's combine those:

```{r combine data}
library(dplyr)
#We use a right join because the 2015 data has more rows
scores10to19 <- dplyr::right_join(scores10.income, scores15, by = "lsoacode")
#Then left join
scores10to19 <- dplyr::left_join(scores10to19, scores19, by = "lsoacode")
```

The overall IMD score is in the 2010 *overall* data but the 2015 *scores breakdown* spreadsheet. We need to combine those.

```{r combine imd scores}
imdscores10to19 <- dplyr::right_join(data10, scores15, by = "lsoacode")
imdscores10to19 <- dplyr::left_join(imdscores10to19, scores19, by = "lsoacode")
colnames(imdscores10to19)[6] <- "imdscore10"
```


## Analysis: how has England changed?

```{r install sqldf}
if(!require("sqldf")){
  install_formats("sqldf")
}
library(sqldf)
```

Let's see what the average is:

```{r average score 1}
sqldf::sqldf("SELECT AVG(imdscore10), AVG(imdscore15), AVG(imdscore19) FROM imdscores10to19")
```
And the median:

```{r average score}
sqldf::sqldf("SELECT MEDIAN(imdscore10), MEDIAN(imdscore15), MEDIAN(imdscore19) FROM imdscores10to19")
```

Very little change there.

How about the other measures?

```{r income change}
sqldf::sqldf("SELECT MEDIAN(incomescore10), MEDIAN(incomescore15), MEDIAN(incomescore19) FROM scores10to19")
```

```{r other change}
sqldf::sqldf("SELECT MEDIAN(employscore15), MEDIAN(employscore19) FROM scores10to19")
sqldf::sqldf("SELECT MEDIAN(IDACIrate15), MEDIAN(IDACIrate19) FROM scores10to19")
sqldf::sqldf("SELECT MEDIAN(IDAOPIrate15), MEDIAN(IDAOPIrate19) FROM scores10to19")

```
